### 2018-06-16 paper/5-implementations.tex
Finish S5.1

We provide an implementation of \Make using our framework in
Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
on a topological sort (see~\S\ref{sec-parallelism} for parallel \Make). For
each key, it builds it if it is older than any of its dependencies. We capture
the persistent build information that \Make stores by a pair
\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
modification times}. The top-level definition
\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
of the scheduler-rebuilder decomposition, illustrated in Table~\ref{tab-build-systems}.

Setting aside the explicit manipulation of file modification times, which in
reality is taken care of by the file system, the function \hs{make} captures
the essence of \Make in a clear and precise manner.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks key = execState $ forM_ chain $ \k -> case tasks k of
    Nothing   -> return ()
    Just task -> do
        value <- gets (getValue k)
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder k value task
            newFetch :: k -> StateT i (State (Store i k v)) v
            newFetch = lift . gets . getValue
        info <- gets getInfo
        (newValue, newInfo) <- runStateT (run newTask newFetch) info
        modify $ putInfo newInfo . putValue k newValue
  where
    deps k = case tasks k of { Nothing -> []; Just task -> dependencies task }
    chain  = case topSort (reachable deps key) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty
    then return value
    else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{topological} calls \hs{process} on every \hs{key} in
a topological order, providing the list of its dependencies \hs{deps} and the
action \hs{act} to compute the resulting value if it needs to be rebuilt. To
determine if the \hs{key} is \hs{dirty}, \hs{process} compares its modification
time with those of its dependencies. If the \hs{key} needs to be rebuilt, the
action \hs{act} is executed and the obtained result is stored, along with an
updated file modification timestamp.

The implementation of \hs{topological} encodes the dependency strategy that
\Make has chosen to use. The \hs{where} clause corresponds to the pre-processing
stage, which uses the function \hs{dependencies}, defined in~\S\ref{sec-deps},
to extract static dependencies from the provided applicative \hs{task}. We
compute the linear processing \hs{chain} by taking the keys \hs{reachable}
from \hs{key} via dependencies, and performing the topological sort of the
result. We omit implementation of textbook graph algorithms
\hs{reachable} and \hs{topSort}, e.g.
see~\cite{cormen2001introduction}.

The \hs{chain} is processed in the \hs{State} monad, with each non-input key
\hs{k} in the chain passed to the provided \hs{process} function, along with
\hs{k}'s dependencies and the action \hs{act}, which when executed recomputes
the \hs{k}'s value by fetching its dependencies from the store.

Note that \hs{dependencies} is only defined for applicative tasks, which is what
restricts \Make to static dependencies, as reflected in the
type~\hs{Build}~\hs{Applicative}. Moreover, any other build system following
the same \hs{topological} approach will also inherit the same restriction.

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty
    then return value
    else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        value <- gets (getValue key)
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key value task
            newFetch :: k -> StateT i (State (Store i k v)) v
            newFetch = lift . gets . getValue
        info <- gets getInfo
        (newValue, newInfo) <- runStateT (run newTask newFetch) info
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

An implementation of \Make using our framework is shown in
Fig.~\ref{fig-make-implementation}. The top-level definition
\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
of the scheduler-rebuilder decomposition in Table~\ref{tab-build-systems}. Below
we describe both components, starting with \hs{modTimeRebuilder}.

We capture persistent build information that \Make stores by the pair
\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
modification times}. The rebuilder receives three arguments: a \hs{key}, its
current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild the
\hs{key} if necessary. The rebuilder returns a modified monadic task of type
\hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
access to the build information and uses it to decide whether to return the
current \hs{value} as is or rebuild the \hs{key} if it is \hs{dirty}, i.e.
if it is older than any of its dependencies\footnote{Since the \hs{task} is
applicative we can statically find its dependencies using the function
\hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
(in which case \hs{modTime} has no entry for it). Note that in our model we
update the current time and file modification time whenever the \hs{key} is
rebuilt, which in reality is taken care of by the file system.

\Make's scheduler processes keys in a linear \hs{order} based on a
\hs{topological} sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
written to the store. The computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
build system (\S\ref{sec-general-build}), but we also need to use an additional
\hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}.

The \hs{where} clause corresponds to the pre-processing stage, which uses the
function \hs{dependencies}, defined in~\S\ref{sec-deps}, to extract static
dependencies from the provided applicative \hs{task}. We compute the linear
processing \hs{order} by constructing the graph of keys \hs{reachable} from the
\hs{target} via dependencies, and performing the topological sort of the result.
We omit implementation of textbook graph algorithms
\hs{reachable} and \hs{topSort}, e.g. see~\cite{cormen2001introduction}.

Note that \hs{dependencies} is only defined for applicative tasks, which is what
restricts \Make to static dependencies, as reflected in the
type~\hs{Build}~\hs{Applicative}. Moreover, any other build system using
the same \hs{topological} scheduler will also inherit the same restriction.
